Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material

### 2019-05-24 paper/supplementary.tex
hide francois last theorem because does not render, feel free to fix
second theorem

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage[]{neurips_2019}

@@

>>>>

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage[]{neurips_2019}
\newtheorem{theorem}{Theorem}

@@

\clearpage
 \newpage
 \begin{theorem}[B2B consistency - general case]

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem}
\begin{proof}

 Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
 \begin{align*}
    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
     \label{eq:doublenorm}
\end{align*}
\begin{align*}
    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
    &= EF \hat G
\end{align*}

Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero.

 Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$). Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore, 
 $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$

But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$.

Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$.

Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
\end{proof}

So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.

Assuming, without loss of generality, that the active features are the $k$ first, we have

$X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right)$, $N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right)$ and

$\hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right)$

Because $\hat H$ is built from covariance matrices of $X$ and $N$, all its diagonal elements are positive, with only the k first different from zero. Their average value is $\frac{1}{k} Tr(\hat H)$. When noise tends to zero, they approach one, and $Diag(\hat H) = E$.

In any case, positive diagonal elements will be bounded by the maximum and minimum value of $\frac{\sigma_X}{\sigma_X + \sigma_N}$ over all eigenvalues of the covariance matrices of $XE$ and $NE$, that is, denoting $\sigma_{X1_1}$, $\sigma_{X1_k}$, $\sigma_{N1_1}$ and $\sigma_{N1_k}$ the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$, all positive diagonal values of $\hat H$ are bounded by $\frac{\sigma_{X1_1}}{\sigma_{X1_1} +\sigma_{N1_k}}$ and $\frac{\sigma_{X1_k}}{\sigma_{X1_k} +\sigma_{N1_1}}$.

On average, and since $X$ and $N$ are uncorrelated, the average value of positive diagonal terms of $\hat H$ should be close to $\frac{Var(X)}{Var(X)+Var(N)}$
